Avoiding tipping points in fisheries management through Gaussian Process Dynamic Programming

### 2014-11-20 manuscripts/manuscript.Rmd
minor typos fixed

Decision making under uncertainty is a ubiquitous challenge in the
management of human intervention in natural resources and conservation.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  Over the last
four decades (beginning with
@Clark1976, @Clark2009 and @Walters1978)
dynamic optimization methods, particularly Stochastic Dynamic Programming
(SDP), have become increasingly important as a means of
understanding how to manage human intervention into natural systems.
Simultaneously, there has been increasing recognition of the importance of
multiple steady states or `tipping points' [@Scheffer2001; .  @Scheffer2009;
@Polasky2011] in ecological systems.

We develop a novel approach to address these concerns in the context of
fisheries; although the challenges and methods are germane to
other problems of conservation or natural resource exploitation.  Economic value and ecological concern have made
marine fisheries the crucible for much of the founding work on management
under uncertainty [@Gordon1954;
@Clark1976; -@Clark2009; @Reed1979; @May1979; @Ludwig1982].

@@

Tipping points in ecological dynamics [@Scheffer2001;
@Polasky2011] highlight this problem because 
precise models are not available and data are
limited such as around  high
stock levels or an otherwise desirable state. With perfect information
, one would know just how far a system could
be pushed before crossing the tipping point, and management would be
simple.  But we face imperfect models and limited data and, with tipping points,  
even small errors can have
very large consequences, as we shall illustrate later.  Because intervention may be too late once a 
tipping point has been crossed (but see @Hughes2013), management
is often concerned with avoiding 
tipping points before any data about them are available.

The dual concerns of model uncertainty and incomplete data 
create a substantial challenge to existing decision-theoretic approaches
[@Brozovic2011].   We illustrate how  Stochastic Dynamic Programming (SDP) 
[@Mangel1988; @Marescot2013] can be implemented using a
Bayesian Non-Parametric (BNP) model of population dynamics [@Munch2005a].
The BNP method has two distinct advantages 
.  First, using a BNP model sidesteps the need for an accurate
model-based description of the system dynamics.  Second, a BNP model
better reflects uncertainty when extrapolating beyond the observed
data.  This is crucial to providing  robust
decision-making when the correct model is not known (as is almost always true).
[We use _robust_ to characterize approaches that provide nearly optimal solutions
without being sensitive to the choice of the (unknown) underlying model.]

@@

[^1]: We abbreviate Gaussian Process as GP, which refers to 
the statistical model we use to approximate the population dynamics
, and we use the
term Gaussian Process Dynamic Programming [GPDP,], to refer to the _use_ of a
GP as the underlying process model when solving a Dynamic Programming equation
.  Hence we will refer to the models as: GP, Ricker, Allen, etc, and the novel method we put forward here as GPDP.

@@

Dynamic optimization  requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function.
For simplicity, we consider only a single state variable, which allows us to
focus on one-dimensional models of population dynamics 
for the comparison of parametric models and the GP.
This is a best-case scenario for the parametric models because we
simulate underlying dynamics from one of the three parametric models
, whereas in the natural world we never know the "true" model.
In addition, by choosing one-dimensional models with just a few parameters, we
limit the chance that poor performance will be due to inability to
estimate parameters accurately, something that becomes a more severe problem
for higher-dimensional parametric models. Finally, the parametric models
we consider are commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable states.

@@

When the parameters governing the dynamics
are also uncertain (the case of parametric uncertainty), the expectation in Eqn 2 involves averaging over the
posterior distribution for the parameters. That is,

@@

We inferred posterior distributions for the parameters of each model by Gibbs sampling (@Gelman2003 implemented in R
[@RTeam] using `jags`, [@R2jags]).  We choose uniform priors
for all parameters (See appendix Tables S1-S3; R code provided). We
show one-step-ahead predictions of these model fits in Figure 1.
We tested each chain for Gelman-Rubin convergence and results were
robust to longer runs.  For each simulation we also applied several
commonly used model selection criteria (AIC, BIC, DIC, see @Burnham2002)
to identify the best fitting model.

@@

\begin{equation}
E = K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I}_n \right)^{-1} \mathbf{Y}_{\textrm{obs}}
\end{equation}

@@

\begin{equation}
C = K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm p}) - K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I} \right)^{-1} K(\mathbf{X}_{\textrm{obs}}, \mathbf{X}_{\textrm p})
\end{equation}

Here $\mathbb{I}_n is the $n$ by $n$ identity matrix (i.e. a matrix with ones down the diagonal and zeros elsewhere) and $K$ is the ‘covariance kernel.’  The covariance kernel controls how much influence one observation has on another.  In the present application we use the squared-exponential kernel which, when evaluated over a pair of vectors, say $x$ and $y$,  generates a covariance matrix whose $i,j$th element is given by

@@

We  derive the harvest policy from the
estimated GP by inserting it into a SDP algorithm. Given the GP posteriors, we construct the transition
matrix representing the probability of going to each state $X(t+1)$ 
given any current state $x(t)$ and any harvest $h(t)$ (See the function
`gp_transition_matrix()` in the provided R package). Given this transition
matrix, we use the same value iteration algorithm as in the parametric
case to determine the optimal policy.
In doing so, 
the uncertainty in the future state under the GP, $X(t+1)$,
includes both process uncertainty (based on the estimation of $\sigma$) 
and structural uncertainty of the posterior collection of curves.
Unlike the case of a parametric model, which is a single curve given the parameters, 
the GP remains a distribution of curves.

@@

All of the models fit the observed data rather closely and with relatively small uncertainty. In Figure 1, we show the posterior predictive curves.  The  training data of stock sizes observed over time are points, overlaid with the step-ahead predictions of each estimated model using the parameters sampled from their posterior distributions. Compared to the true model most estimates appear to over-fit, predicting patterns that are actually due purely to stochasticity.  Model selection criteria (Table 1) penalize more complex models and show a preference for the simpler Ricker model over the  models with alternative stable states (Allen and Myers).  Details on MCMC estimates for each model, traces, and posterior distributions can be found in the appendix.

@@

In Figure 4, we show the consequences of managing 100 replicate
realizations of the simulated fishery under policies derived from each model
.  The structurally
correct model under-harvests, leaving the stock to vary around its
unfished optimum.  The structurally incorrect Ricker model over-harvests
the population past the tipping point consistently, resulting in the
immediate crash of the stock and thus leads to minimal long-term catch.

@@

Changing the intensity of the stochasticity or the distance between stable and unstable steady states
does not impact the performance of the GP relative to the optimal solution
obtained from the true model and true parameters.  The parametric models are
more sensitive to this difference.  Large values of $\sigma$ relative to the distance between the stable and unstable
point increases the chance of a stochastic transition below the tipping point.  More precisely,
if we let $L$ be the distance between the stable and unstable steady states, then the probability that fluctuations drive the population across the unstable steady state scales as
$\exp\left(\frac{L^2}{\sigma^2}\right)$

(see @Gardiner2009 or @Mangel2006 for the derivation).
Thus, the impact of using a model that  underestimates the risk of harvesting
beyond the critical point is considerable, since this such a situation
occurs more often.  Conversely, with large enough distance between the
optimal escapement and unstable steady state relative to $\sigma$, the chance
of a transition becomes vanishingly small and all models can be estimated 
near-optimally. Models that underestimate the cost incurred by population
sizes fluctuating significantly below the optimal escapement level will
not perform poorly as long as those fluctuations are sufficiently small.

The GPDP is only weakly influenced by increasing stochasticity or increasing
Allee effects over much of the range (Figure 6). Larger $\sigma$ or higher Allee
levels make even the optimal solution without any model or parameter uncertainty
unable to harvest the population effectively (e.g. the stochasticity is large enough to violate the
self-sustaining criterion of @Reed1979).

```{r figure_6, fig.width=6, fig.height=3, dependson=c("sensitivity-trends"), fig.cap="The effect of increasing noise or decreasing Allee threshold levels on the net present value of the fishery when managed under the GPDP, relative to managing under the true model (with known parameters).  Other than the focal parameter (stochasticity, Allee threshold), other parameters are held fixed as above to illustrate this effect.", cache=FALSE}
plot_sigmas <- ggplot(vary_sigma, aes(noise, value)) + 
  geom_point(size=2) + ylab("Net present value") + 
  xlab("Level of growth stochasticity")
plot_allees <- ggplot(vary_allee, aes(allee, value)) + 
  geom_point(size=2)+ ylab("Net present value") + 
  xlab("Allee threshold stock size")
multiplot(plot_sigmas, plot_allees, cols=2) 
write.csv(vary_sigma, "components/data/figure6a.csv")
write.csv(vary_allee, "components/data/figure6b.csv")
```

>>>>

Decision making under uncertainty is a ubiquitous challenge in the
management of human intervention in natural resources and conservation.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  Over the last
four decades (beginning with
@Clark1976, @Clark2009 and @Walters1978)
dynamic optimization methods, particularly Stochastic Dynamic Programming
(SDP), have become increasingly important as a means of
understanding how to manage human intervention into natural systems.
Simultaneously, there has been increasing recognition of the importance of
multiple steady states or `tipping points' [@Scheffer2001; @Scheffer2009;
@Polasky2011] in ecological systems.

We develop a novel approach to address these concerns in the context of
fisheries; although the challenges and methods are germane to
other problems of conservation or natural resource exploitation.  Economic value and ecological concern have made
marine fisheries the crucible for much of the founding work on management
under uncertainty [@Gordon1954; @Clark1976; @Clark2009; @Reed1979; @May1979; @Ludwig1982].

@@

Tipping points in ecological dynamics [@Scheffer2001;
@Polasky2011] highlight this problem because 
precise models are not available and data are
limited such as around  high
stock levels or an otherwise desirable state. With perfect information, 
one would know just how far a system could
be pushed before crossing the tipping point, and management would be
simple.  But we face imperfect models and limited data and, with tipping points,  
even small errors can have
very large consequences, as we shall illustrate later.  Because intervention may be too late once a 
tipping point has been crossed (but see @Hughes2013), management
is often concerned with avoiding 
tipping points before any data about them are available.

The dual concerns of model uncertainty and incomplete data 
create a substantial challenge to existing decision-theoretic approaches
[@Brozovic2011].   We illustrate how  Stochastic Dynamic Programming (SDP) 
[@Mangel1988; @Marescot2013] can be implemented using a
Bayesian Non-Parametric (BNP) model of population dynamics [@Munch2005a].
The BNP method has two distinct advantages. First, using a BNP model sidesteps the need for an accurate
model-based description of the system dynamics.  Second, a BNP model
better reflects uncertainty when extrapolating beyond the observed
data.  This is crucial to providing  robust
decision-making when the correct model is not known (as is almost always true).
[We use _robust_ to characterize approaches that provide nearly optimal solutions
without being sensitive to the choice of the (unknown) underlying model.]

@@

[^1]: We abbreviate Gaussian Process as GP, which refers to 
the statistical model we use to approximate the population dynamics, and we use the
term Gaussian Process Dynamic Programming [GPDP,], to refer to the _use_ of a
GP as the underlying process model when solving a Dynamic Programming equation.
Hence we will refer to the models as: GP, Ricker, Allen, etc, and the novel method we put forward here as GPDP.

@@

Dynamic optimization  equires characterizing the dynamics of a state
variable (or variables), a control action, and a value function.
For simplicity, we consider only a single state variable.
This is a best-case scenario for the parametric models because we
simulate underlying dynamics from one of the three parametric models
, whereas in the natural world we never know the "true" model.
In addition, by choosing one-dimensional models with just a few parameters, we
limit the chance that poor performance will be due to inability to
estimate parameters accurately, something that becomes a more severe problem
for higher-dimensional parametric models. Finally, the parametric models
we consider are commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable states.

@@

When the parameters governing the dynamics
are also uncertain, we take the expectation over the
posterior distribution for the parameters:

@@

We inferred posterior distributions for the parameters of each model by Gibbs sampling (@Gelman2003 implemented in R
[@RTeam] using `jags`, [@R2jags]).  We choose uniform priors
for all parameters of the parametric models (See appendix Tables S1-S3; R code provided). We
show one-step-ahead predictions of these model fits in Figure 1.
We tested each chain for Gelman-Rubin convergence and results were
robust to longer runs.  For each simulation we also applied several
commonly used model selection criteria (AIC, BIC, DIC, see @Burnham2002)
to identify the best fitting model.

@@

\begin{equation}
E = K(X_p, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I}_n \right)^{-1} \mathbf{Y}_{\textrm{obs}}
\end{equation}

@@

\begin{equation}
C = K(X_p, X_p) - K(X_p, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I} \right)^{-1} K(\mathbf{X}_{\textrm{obs}}, X_p)
\end{equation}

Here $\mathbb{I}_n$ is the $n$ by $n$ identity matrix (i.e. a matrix with ones down the diagonal and zeros elsewhere) and $K$ is the ‘covariance kernel.’  The covariance kernel controls how much influence one observation has on another.  In the present application we use the squared-exponential kernel which, when evaluated over a pair of vectors, say $\mathbf{x}$ and $\mathbf{y}$,  generates a covariance matrix whose $i,j$th element is given by

@@

We  derive the harvest policy from the
estimated GP by inserting it into a SDP algorithm. Given the GP posteriors, we construct the transition
matrix representing the probability of going to each state $X(t+1)$ 
given any current state $x(t)$ and any harvest $h(t)$ (See the function
`gp_transition_matrix()` in the provided R package). Given this transition
matrix, we use the same value iteration algorithm as in the parametric
case to determine the optimal policy. In doing so, 
the uncertainty in the future state under the GP, $X(t+1)$,
includes both process uncertainty (based on the estimation of $\sigma$) 
and structural uncertainty of the posterior collection of curves.

@@

All of the models fit the observed data rather closely and with relatively small uncertainty. In Figure 1, we show the posterior predictive curves.  The  training data of stock sizes observed over time are points, overlaid with the step-ahead predictions of each estimated model using the parameters sampled from their posterior distributions. Compared to the true model most estimates appear to over-fit, predicting patterns that are actually due purely to stochasticity.  Model selection criteria (Table 1) penalize more complex models and show a preference for the simpler Ricker model over the  models with alternative stable states (Allen and Myers). Supplement provides details on the model estimates.

@@

In Figure 4, we show the consequences of managing 100 replicate
realizations of the simulated fishery under policies derived from each model.  The structurally
correct model under-harvests, leaving the stock to vary around its
unfished optimum.  The structurally incorrect Ricker model over-harvests
the population past the tipping point consistently, resulting in the
immediate crash of the stock and thus leads to minimal long-term catch.

@@

The GPDP is only weakly influenced by increasing stochasticity or increasing
Allee effects over much of the range (Figure S2). Larger $\sigma$ or higher Allee
levels make even the optimal solution without any model or parameter uncertainty
unable to harvest the population effectively (e.g. the stochasticity is large enough to violate the
self-sustaining criterion of @Reed1979).

### 2014-11-20 manuscripts/manuscript.tex
minor typos fixed

Tipping points in ecological dynamics (Scheffer et al. 2001, Polasky et
al. 2011) highlight this problem because precise models are not
available and data are limited such as around high stock levels or an
otherwise desirable state. With perfect information , one would know
just how far a system could be pushed before crossing the tipping point,
and management would be simple. But we face imperfect models and limited
data and, with tipping points,\\even small errors can have very large
consequences, as we shall illustrate later. Because intervention may be
too late once a tipping point has been crossed (but see Hughes et al.
(2013)), management is often concerned with avoiding tipping points
before any data about them are available.

The dual concerns of model uncertainty and incomplete data create a
substantial challenge to existing decision-theoretic approaches
(Brozović and Schlenker 2011). We illustrate how Stochastic Dynamic
Programming (SDP) (Mangel and Clark 1988, Marescot et al. 2013) can be
implemented using a Bayesian Non-Parametric (BNP) model of population
dynamics (Munch et al. 2005a). The BNP method has two distinct
advantages . First, using a BNP model sidesteps the need for an accurate
model-based description of the system dynamics. Second, a BNP model
better reflects uncertainty when extrapolating beyond the observed data.
This is crucial to providing robust decision-making when the correct
model is not known (as is almost always true). {[}We use \emph{robust}
to characterize approaches that provide nearly optimal solutions without
being sensitive to the choice of the (unknown) underlying model.{]}

@@

We first describe the requirements of dynamic optimization for the
management of human intervention in natural resource systems. After that
we describe three parametric models for population dynamics and the
Gaussian Process (GP)\footnote{We abbreviate Gaussian Process as GP,
  which refers to the statistical model we use to approximate the
  population dynamics , and we use the term Gaussian Process Dynamic
  Programming {[}GPDP,{]}, to refer to the \emph{use} of a GP as the
  underlying process model when solving a Dynamic Programming equation .
  Hence we will refer to the models as: GP, Ricker, Allen, etc, and the
  novel method we put forward here as GPDP.} description of population
dynamics.

@@

Dynamic optimization requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function. For
simplicity, we consider only a single state variable, which allows us to
focus on one-dimensional models of population dynamics for the
comparison of parametric models and the GP. This is a best-case scenario
for the parametric models because we simulate underlying dynamics from
one of the three parametric models , whereas in the natural world we
never know the ``true'' model. In addition, by choosing one-dimensional
models with just a few parameters, we limit the chance that poor
performance will be due to inability to estimate parameters accurately,
something that becomes a more severe problem for higher-dimensional
parametric models. Finally, the parametric models we consider are
commonly used in modeling stock-recruitment dynamics or to model sudden
transitions between alternative stable states.

@@

We inferred posterior distributions for the parameters of each model by
Gibbs sampling (Gelman et al. (2003) implemented in R (R Core Team 2013)
using \texttt{jags}, (Su and Masanao Yajima 2013)). We choose uniform
priors for all parameters (See appendix Tables S1-S3; R code provided).
We show one-step-ahead predictions of these model fits in Figure 1. We
tested each chain for Gelman-Rubin convergence and results were robust
to longer runs. For each simulation we also applied several commonly
used model selection criteria (AIC, BIC, DIC, see Burnham and Anderson
(2002)) to identify the best fitting model.

@@

\begin{equation}
C = K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm p}) - K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I} \right)^{-1} K(\mathbf{X}_{\textrm{obs}}, \mathbf{X}_{\textrm p})
\end{equation}

Here $\mathbb{I}_n is the $n\$ by $n$ identity matrix (i.e.~a matrix
with ones down the diagonal and zeros elsewhere) and $K$ is the
`covariance kernel.' The covariance kernel controls how much influence
one observation has on another. In the present application we use the
squared-exponential kernel which, when evaluated over a pair of vectors,
say $x$ and $y$, generates a covariance matrix whose $i,j$th element is
given by

@@

We derive the harvest policy from the estimated GP by inserting it into
a SDP algorithm. Given the GP posteriors, we construct the transition
matrix representing the probability of going to each state $X(t+1)$
given any current state $x(t)$ and any harvest $h(t)$ (See the function
\texttt{gp\_transition\_matrix()} in the provided R package). Given this
transition matrix, we use the same value iteration algorithm as in the
parametric case to determine the optimal policy. In doing so, the
uncertainty in the future state under the GP, $X(t+1)$, includes both
process uncertainty (based on the estimation of $\sigma$) and structural
uncertainty of the posterior collection of curves. Unlike the case of a
parametric model, which is a single curve given the parameters, the GP
remains a distribution of curves.

@@

Despite the similarities in model fits to the observed data, the
policies inferred under each model differ widely (Figure 3). Policies
are shown in terms of target escapement, $S(t) = x_t - h$. Under models
such as this a constant escapement policy is expected to be optimal
(Reed 1979), whereby population levels below a certain size $S$ are
unharvested, while above that size the harvest strategy aims to return
the population to $S$. Whenever a model predicts that the population
will not persist below a certain threshold, the optimal solution is to
harvest the entire population immediately, resulting in an escapement
$S=0$, as seen in the true (correct form, exact parameters) model, the
Allen model (correct form, estimated parameters) and the GP.\\Only the
structurally correct model (Allen model) and the GP produce policies
close to the true optimum policy.

@@

Changing the intensity of the stochasticity or the distance between
stable and unstable steady states does not impact the performance of the
GP relative to the optimal solution obtained from the true model and
true parameters. The parametric models are more sensitive to this
difference. Large values of $\sigma$ relative to the distance between
the stable and unstable point increases the chance of a stochastic
transition below the tipping point. More precisely, if we let $L$ be the
distance between the stable and unstable steady states, then the
probability that fluctuations drive the population across the unstable
steady state scales as $\exp\left(\frac{L^2}{\sigma^2}\right)$

(see Gardiner (2009) or Mangel (2006) for the derivation). Thus, the
impact of using a model that underestimates the risk of harvesting
beyond the critical point is considerable, since this such a situation
occurs more often. Conversely, with large enough distance between the
optimal escapement and unstable steady state relative to $\sigma$, the
chance of a transition becomes vanishingly small and all models can be
estimated near-optimally. Models that underestimate the cost incurred by
population sizes fluctuating significantly below the optimal escapement
level will not perform poorly as long as those fluctuations are
sufficiently small.

The GPDP is only weakly influenced by increasing stochasticity or
increasing Allee effects over much of the range (Figure 6). Larger
$\sigma$ or higher Allee levels make even the optimal solution without
any model or parameter uncertainty unable to harvest the population
effectively (e.g.~the stochasticity is large enough to violate the
self-sustaining criterion of Reed (1979)).

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_6-1.pdf}
\caption{The effect of increasing noise or decreasing Allee threshold
levels on the net present value of the fishery when managed under the
GPDP, relative to managing under the true model (with known parameters).
Other than the focal parameter (stochasticity, Allee threshold), other
parameters are held fixed as above to illustrate this effect.}
\end{figure}

@@

Gardiner, C. 2009. Stochastic Methods: A Handbook for the Natural and
Social Sciences (Springer Series in Synergetics). Page 447. Springer.

@@

Mangel, M. 2006. The Theoretical Biologist's Toolbox: Quantitative
Methods for Ecology and Evolutionary Biology. Cambridge University
Press.

>>>>

Tipping points in ecological dynamics (Scheffer et al. 2001, Polasky et
al. 2011) highlight this problem because precise models are not
available and data are limited such as around high stock levels or an
otherwise desirable state. With perfect information, one would know just
how far a system could be pushed before crossing the tipping point, and
management would be simple. But we face imperfect models and limited
data and, with tipping points,\\even small errors can have very large
consequences, as we shall illustrate later. Because intervention may be
too late once a tipping point has been crossed (but see Hughes et al.
(2013)), management is often concerned with avoiding tipping points
before any data about them are available.

The dual concerns of model uncertainty and incomplete data create a
substantial challenge to existing decision-theoretic approaches
(Brozović and Schlenker 2011). We illustrate how Stochastic Dynamic
Programming (SDP) (Mangel and Clark 1988, Marescot et al. 2013) can be
implemented using a Bayesian Non-Parametric (BNP) model of population
dynamics (Munch et al. 2005a). The BNP method has two distinct
advantages. First, using a BNP model sidesteps the need for an accurate
model-based description of the system dynamics. Second, a BNP model
better reflects uncertainty when extrapolating beyond the observed data.
This is crucial to providing robust decision-making when the correct
model is not known (as is almost always true). {[}We use \emph{robust}
to characterize approaches that provide nearly optimal solutions without
being sensitive to the choice of the (unknown) underlying model.{]}

@@

We first describe the requirements of dynamic optimization for the
management of human intervention in natural resource systems. After that
we describe three parametric models for population dynamics and the
Gaussian Process (GP)\footnote{We abbreviate Gaussian Process as GP,
  which refers to the statistical model we use to approximate the
  population dynamics, and we use the term Gaussian Process Dynamic
  Programming {[}GPDP,{]}, to refer to the \emph{use} of a GP as the
  underlying process model when solving a Dynamic Programming equation.
  Hence we will refer to the models as: GP, Ricker, Allen, etc, and the
  novel method we put forward here as GPDP.} description of population
dynamics.

@@

Dynamic optimization equires characterizing the dynamics of a state
variable (or variables), a control action, and a value function. For
simplicity, we consider only a single state variable. This is a
best-case scenario for the parametric models because we simulate
underlying dynamics from one of the three parametric models , whereas in
the natural world we never know the ``true'' model. In addition, by
choosing one-dimensional models with just a few parameters, we limit the
chance that poor performance will be due to inability to estimate
parameters accurately, something that becomes a more severe problem for
higher-dimensional parametric models. Finally, the parametric models we
consider are commonly used in modeling stock-recruitment dynamics or to
model sudden transitions between alternative stable states.

@@

We inferred posterior distributions for the parameters of each model by
Gibbs sampling (Gelman et al. (2003) implemented in R (R Core Team 2013)
using \texttt{jags}, (Su and Masanao Yajima 2013)). We choose uniform
priors for all parameters of the parametric models (See appendix Tables
S1-S3; R code provided). We show one-step-ahead predictions of these
model fits in Figure 1. We tested each chain for Gelman-Rubin
convergence and results were robust to longer runs. For each simulation
we also applied several commonly used model selection criteria (AIC,
BIC, DIC, see Burnham and Anderson (2002)) to identify the best fitting
model.

@@

\begin{equation}
C = K(X_p, X_p) - K(X_p, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I} \right)^{-1} K(\mathbf{X}_{\textrm{obs}}, X_p)
\end{equation}

Here $\mathbb{I}_n$ is the $n$ by $n$ identity matrix (i.e.~a matrix
with ones down the diagonal and zeros elsewhere) and $K$ is the
`covariance kernel.' The covariance kernel controls how much influence
one observation has on another. In the present application we use the
squared-exponential kernel which, when evaluated over a pair of vectors,
say $\mathbf{x}$ and $\mathbf{y}$, generates a covariance matrix whose
$i,j$th element is given by

@@

We derive the harvest policy from the estimated GP by inserting it into
a SDP algorithm. Given the GP posteriors, we construct the transition
matrix representing the probability of going to each state $X(t+1)$
given any current state $x(t)$ and any harvest $h(t)$ (See the function
\texttt{gp\_transition\_matrix()} in the provided R package). Given this
transition matrix, we use the same value iteration algorithm as in the
parametric case to determine the optimal policy. In doing so, the
uncertainty in the future state under the GP, $X(t+1)$, includes both
process uncertainty (based on the estimation of $\sigma$) and structural
uncertainty of the posterior collection of curves.

@@

Despite the similarities in model fits to the observed data, the
policies inferred under each model differ widely (Figure 3). Policies
are shown in terms of target escapement, $S(t) = x_t - h$. Under models
such as this a constant escapement policy is expected to be optimal
(Reed 1979), whereby population levels below a certain size $S$ are
unharvested, while above that size the harvest strategy aims to return
the population to $S$. Whenever a model predicts that the population
will not persist below a certain threshold, the optimal solution is to
harvest the entire population immediately, resulting in an escapement
$S=0$, as seen in the true (correct form, exact parameters) model, the
Allen model (correct form, estimated parameters) and the GP. Only the
structurally correct model (Allen model) and the GP produce policies
close to the true optimum policy.

@@

The GPDP is only weakly influenced by increasing stochasticity or
increasing Allee effects over much of the range (Figure S2). Larger
$\sigma$ or higher Allee levels make even the optimal solution without
any model or parameter uncertainty unable to harvest the population
effectively (e.g.~the stochasticity is large enough to violate the
self-sustaining criterion of Reed (1979)).

@@

@@
